---
knit: "bookdown::preview_chapter"
---

# Strings

Wrangling strings of characters is something we
will find ourselves doing often as data scientists. R provides a
comprehensive set of tools for handling and processing strings. In
this chapter we review the functionality provided by R for managing
and manipulating strings.

## Packages from the R Library

Packages used in this chapter include
_stringr_,
_stringi_,
_magrittr_, and
_rattle_.

```{r strings:attach_packages, warning=FALSE, message=FALSE}
# Load required packages from local library into R session.

library(stringi)      # The string concat operator %s+%.
library(stringr)      # String manipulation.
library(magrittr)     # Pipelines for data processing: %>% %T>% %<>%.
library(rattle)       # Weather dataset.
``` 

```{r strings:additional_dependent_packages, echo=FALSE, message=FALSE, purl=FALSE}
# These are dependencies that would otherwise be attached as required.

library(xtable)
library(readr)
```

## Concatenate Strings

One of the most basic operations in string
manipulation is the concatenate operation. R provides alternatives
for doing so but a modern favourite is the %s+%
operator.

```{r strings:concatenate}
"abc" %s+% "def" %s+% "ghi"
c("abc", "def", "ghi", "jkl") %s+% c("mno")
c("abc", "def", "ghi", "jkl") %s+% c("mno", "pqr")
c("abc", "def", "ghi", "jkl") %s+% c("mno", "pqr", "stu", "vwx")
```

The tidy function for concatenating strings is
**str_c()**. A _sep=_ can be used to specify a
separator for the concatenated strings.

```{r strings:str_c}
str_c("hello", "world")
str_c("hello", "world", sep=" ")
```

The traditional **cat()** function returns the
concatenation of the supplied strings.  Numeric and other complex
objects are converted into character strings. 

```{r strings:cat}
cat("hello", "world")
cat ("hello", 123, "world")
```

Yet another alternative (and there are many) is the function
**paste()**. Notice that it separates the concatenated
strings with a space.

```{r strings:paste}
paste("hello", "world")
``` 

## Concatenate Strings Special Cases

Each operator/functions treats NULL
differently. Note the convenience for **cat()** to add a
space between the strings, and that **paste()** treats
NULL as a zero length string, and thus there are two spaces between
the words concatenated.

```{r strings:concat_null}
"hello" %s+% NULL %s+% "world"
str_c("hello", NULL, "world")
cat("hello", NULL, "world")
paste("hello", NULL, "world")

``` 

## String Length

The tidy way to get the length of a string is
**str_length()**.

```{r }
str_length("hello world")
str_length(c("hello", "world"))
str_length(NULL)
str_length(NA)
``` 

The function **nchar()** is the traditional approach.

```{r strings:nchar}
nchar("hello world")
nchar(c("hello", "world"))
nchar(NULL)
nchar(NA)
``` 

## Case Conversion

Often during data transformations strings have to
be converted from one case to the other. These simple transformations
can be achieved by **tolower()** and
**toupper()**. The **casefold()** function
can also be used as a wrapper to the two functions.

```{r strings:case}

toupper("String Manipulation")
tolower("String Manipulation")
casefold("String Manipulation")
casefold("String Manipulation", upper=TRUE)
``` 

## Tidy Sub-String Operations

We will find ourselves often wanting to extract or
modify sub-strings within a string. The tidy way to do this is with
the **str_sub()** function. We can specify the
_start=_ and the _end=_ of the string. The indices of
the string start from 1. 

```{r strings:str_sub}
s <- "string manipulation"
str_sub(s, start=3, end=6)
str_sub(s, 3, 6)
```

A negative is used to count from the end of
the string.

```{r }
str_sub(s, 1, -8)
```

Replacing a sub-string with another string is straightforward using
the assignment operator.

```{r }
str_sub(s, 1, -8) <- "stip"
s
```

The function also operates over a vector of strings.

```{r }
v <- c("string", "manipulation", "always", "fascinating")
str_sub(v, -4, -1)
str_sub(v, -4, -1) <- "RING"
v
``` 

## Base Sub-String Operations

The base function **substr()** can be used to extract and
replace parts of a string similar to **str_sub()**. Note
however that it does not handle negative values and that string
replacement only replaces the same length as the replacement string,
without changing the length of the original string.

```{r }
s <- "string manipulation"
substr(s, start=3, stop=6)
substr(s, 3, 6)
substr(s, 1, 12) <- "stip"
s
``` 

The **substring()** function performs similarly though
uses _last=_ rather than _stop=_.

```{r }
s <- "string manipulation"
substring(s, first=3, last=6)
x <- c("abcd", "aabcb", "babcc", "cabcd")
substring(x, 2, 4)
substring(x, 2, 4) <- "AB"
x
```

## Trim and Pad

One of the major challenges of string parsing is
removing and adding whitespaces and wrapping text.

Additional white space can be present on the left, right or both sides
of the word. The **str_trim()** function offers an
effective way to get rid of these whitespaces.

```{r strings:str_trim}
ws <- c(" abc",  "def ", " ghi ")
str_trim(ws)
str_trim(ws, side="left")
str_trim(ws, side="right")
str_trim(ws, side="both")
``` 

Conversely we can also pad a string with additional characters for up
to a specified width using **str_pad()**. The default
padding character is a space but we can override that.

```{r strings:str_pad}
str_pad("abc", width=7)
str_pad("abc", width=7, side="left")
str_pad("abc", width=7, side="right")
str_pad("abc", width=7, side="both", pad="#")
```

## Wrapping and Words

Formatting a text string into a neat paragraph of
defined maximum width is another operation we often find ourselves
wanting. The **str_wrap()** function will do this for us.

```{r strings:str_wrap}
st <- "All the Worlds a stage, All men are merely players"
cat(str_wrap(st, width=25))
``` 

Words of course form the basis for wrapping a sentence.  We may wish
to extract words from a sentence ourselves for further
processing. Here we us **word()** to do so.We specify the
positions of the word to be extracted from the setence. The default
separator value is space.

```{r strings:word}
st <- c("The quick brown fox", "jumps on the brown dog")
word(st, start=1, end=2)
word(st, start=1, end=-2)
``` 

## Pattern Matching with Regular Expressions

One of the most powerful string processing
concepts is the concept of regular expressions. A regular expression
is a sequence of characters that describe a pattern. The concept was
formalized by American mathematician Stephen Cole Kleene. A regular
expression pattern can contain a combination of alphanumeric and
special characters. It is a complex topic and we take an introductory
look at it here to craft regular expressions in R.

An important concept is that of metacharacters which have special
meaning within a regular expression. Unlike other characters that are
used to match themselves, metacharacters have a specific meaning The
following table shows a list of metacharacters used in regular
expressions.

```{r strings:regexp_metacharacters, results="asis", echo=FALSE, message=FALSE}
"tables/metacharacters.csv" %>%
  read_csv() %>%
  xtable()
```

Such metacharacters are used to match different patterns.

```{r strings:regexp_examples}

s <- c("hands", "data", "on", "data$cience", "handsondata$cience", "handson")
grep(pattern="^data", s, value=TRUE)
grep(pattern="on$", s, value=TRUE)
grep(pattern="(nd)..(nd)", s, value=TRUE)
``` 

In order to match a metacharacter in R we need to escap it with
$\backslash\backslash$ (double backslash).

```{r }
grep(pattern="\\$", s, value=TRUE)
``` 


## Regular Expressions: Quantifiers

Quantifiers are used to match repitition of a
pattern within a string. The following table shows a list of
quantifiers.

```{r strings:regexp_quantifiers, results="asis", echo=FALSE, message=FALSE}
"tables/quantifier.tsv" %>%
  read_tsv() %>%
  xtable()
``` 

Some examples will illustrate.

```{r strings:regexp_qualifiers_examples}
s <- c("aaab", "abb", "bc", "abbcd", "bbbc", "abab", "caa")
grep(pattern="ab*b", s, value=TRUE)
grep(pattern="abbc?", s, value=TRUE)
grep(pattern="b{2,}?", s, value=TRUE)
``` 

## Regular Expressions: Character Classes

A character class is a collection of characters
that are in some way grouped together. We enclose the characters to be
grouped within square backets []. The pattern then matches any one of
the characters in the set. For example, the character class [0-9]
matches any of the digits from 0 to 9.

```{r strings_regexp_digits, results="asis", echo=FALSE, message=FALSE}
"tables/characterclass.tsv" %>%
  read_tsv() %>%
  xtable()
``` 

```{r strings:regex_classes}
s <- c("abc12", "@#$", "345", "ABcd")
grep(pattern="[0-9]+", s, value=TRUE)
grep(pattern="[A-Z]+", s, value=TRUE)
grep(pattern="[^@#$]+", s, value=TRUE)
``` 

R also supports the use of POSIX character classes which are
represented within [[]] (double braces).

```{r strings:regex_posix}

grep(pattern="[[:alpha:]]", s, value=TRUE)
grep(pattern="[[:upper:]]", s, value=TRUE)
``` 

## Generate Strings for Testing

It is sometimes very useful to be able to test out
some code using some test data. A simple way to generate test strings
us with **stri_rand_lipsum()**.

```{r strings:generate_random_text, out.lines=NULL}
stri_rand_lipsum(20)
stri_rand_lipsum(2)
sapply(stri_rand_lipsum(10), nchar, USE.NAMES=FALSE)
sapply(stri_rand_lipsum(10), nchar, USE.NAMES=FALSE)
```

The strings generated are of different lengths and each call generates
different strings.

## Read a File as Vector of Strings

There may be occasions where we would like to load
a dataset from a file as strings, one line as a string, returning a
vector of strings. We can achieve using the function
**readLines()**. IN the following example we access the
system file \texttt{weather.csv} that is provided by the
\Sexpr{Rpackage(rattle)} package.

```{r strings:load_sample_data}
dsname <- "weather" # Dataset name.
ftype  <- "csv"     # Source dataset file type.
dsname  %s+%
  "."   %s+%
  ftype %T>%
  print() %>%
  system.file(ftype, ., package="rattle") %>%
  readLines() ->
ds
``` 

A sample of the data.

```{r strings:head_sample_data}
head(ds)
``` 

Find those strings that contain a specific pattern using
**grep()**.

```{r strings:pattern_matching}
grep("ENE", ds)
grep("ENE", ds, value=TRUE)
``` 



# Resources

* [Data Mining with Rattle and R](https://bit.ly/rattle_data_mining);
* [The Essentials of Data Science](https://bit.ly/essentials_data_science);


*  [Handling
    and Processing Strings in R](http://gastonsanchez.com/Handling_and_Processing_Strings_in_R.pdf), a freely available ebook by Gaston
  Sanchez from 2013.
* <http://www.rexamine.com/2013/04/properly-internationalized-regular-expressions-in-r/>





