---
knit: "bookdown::preview_chapter"
---

# Strings Manipulation

In this module we introduce to tools available in R for handling and
processing strings.  The required packages for this module include:

```{r results='hide', message=FALSE}
library(rattle)  # The weather dataset.
library(stringr) # Pre-eminent package for string handling.
```

## String Concatenation {-}

Let us start with the simplest of string operations - concatenation
two strings.

### cat {-}

The **cat()** function concatenates objects and could also print them
to screen or to a file. By default it converts even numeric and other
complex objects into character type and then concatenates.

```{r Example 1: cat() function, , eval=TRUE}
cat("hello", "world", sep="\t")

# Using numeric values with cat().
x <- 123
cat ("hello", x, sep="\t")
```

###paste {-}

Alternatively we can use the **paste()** function to concatenate and
print the values to screen.

```{r Example 2: paste() function}
paste("hello", "world", sep="\t") 
```

The major disadvantage of the paste function is that it does not
handle null characters effectively. It converts the NULL character as
a space instead of treating it appropriately. **WHAT IS APPROPRIATE?**

```{r Example 3: paste() function with null values}
paste('hello', NULL, 'world')
```

###str_c {-}

The **str_c()** function is similar to the paste() function but the
default separator is white space and it ignores NULL characters.

```{r Example 4: str_c() function handling null values}
str_c('hello', NULL, 'world') # Using str_c with null characters.
```

## String Length  {-}

###nchar {-}

The **nchar()** function in the base package is used to measure the
typical length of the string. One major issues is it's inability to
handle factor values robustly.

```{r Example 5: nchar() function, error=TRUE}
nchar('hello world') # Usage of the nchar function.

factor.Example <- factor(c(1, 1, 0, 0), labels=c("success", "fail"))
nchar(factor.Example) # Using nchar function with factors.

```

###str_length {-}

Alternatively **str_length()** package could also be used to measure
string lengths. In comparisson the str\_length() package handles NA
characters more accurately as nchar(NA) returns 2 while str\_length()
returns NA and also handles factor values appropriately.

```{r Example 6: str_length function, error=TRUE}

str_length('hello world') # Usage of the str_length function.
str_length(factor_example) # Using str_length with factor variables.
```

## Case Conversion {-}

Often during data transformations strings have to be converted from
one case to the other. These simple transformations could be achieved
by **tolower()** and **toupper()** functions. The **casefolding()**
function could also be used as a wrapper to the two functions.

```{r Example 7: case conversions, message=FALSE}

toupper('string manipulation') # Converting text to upper case.
tolower('STRING MANIPULATION') # Converting text to lower case.
casefold('string manipulation', upper=TRUE) # Usage of the casefold function.
```

## Substring Operation {-}

Finding substrings are one of the most common string manipulation
operations.

###**substr** {.unnumbered}

The **substr()** could be used to extract, replace parts of the
string.

```{r Example 8: substr() function}

substr('string manipulation', 3, 6) # Usage of substr function.
s <- 'string manipulation'
substr(s, 3, 6) <- 'RING'   # Replacing strings with the substr function.
s
```

###**substring** {.unnumbered}
The **substring()** functions performs the same operations on a character vector. 

```{r Example 9: substring function}
x <- c('abcd', 'aabcb', 'babcc', 'cabcd') # Extracting characters using the substring function   
substring(x, 2, last=4)
substring(x, 2, last=4) <- 'AB' # Replacing strings with the substring function.
x
```

###str_sub {.unnumbered}

The stringr package offers **str_sub()** which is a equivalent of substring(). The str\_sub() function handles negative values even more robustly than the substring() function.

```{r Example 10: str_sub function}
y <- c("string", "manipulation", "always", "fascinating")

substring(y, first=-4, last=-1) # Using the substring functions with negative indices.
str_sub(y ,  start=-4, end=-1)  # Using the str_sub function to handle negative indices.
str_sub(y, start=-4, end=-1) <- 'RING' # String replacement using the str_sub function.
y
```

## Simple String Operations {-} 

One of the major challenges of string parsing is handline additional whitespaces in words. Often additional widespaces are present on the left, right or both sides of the word. 

###String Trimming {.unnumbered}

The **str_trim** function offers an effective way to get rid of these whitespaces.

```{r Example 11: string trimming}
whitespace.vector <- c('  abc', 'def   ', '     ghi       ')
str_trim(whitespace.vector, side='left') # Trimming white spaces on the left side of the string.
str_trim(whitespace.vector, side='right') # Trimming white spaces on the right side of the string.
str_trim(whitespace.vector, side='both') # Trimming white spaces on both sides of the string.
```

### String Padding {.unnumbered}

Conversely we could also pad a string with additional characters for a defined width using the **str_pad()**
function. The default padding character is a space.

```{r Example 12: string padding}
str_pad('abc', width=7, side="left") # Padding characters to the left side of the string.
str_pad('abc', width=7, side="right") # Padding characters to the right side of the string.
str_pad('abc', width=7, side="both", pad="#") # Padding other characters to both sides of a string.
```

### String Wrapping {.unnumbered}

Sometimes text have to be manipulated to neat paragraphs of defined width. The **str_wrap()** function could be used to format the text into defined paragraphs of specific width.

```{r Example 13: string wrapping}

some.text <- 'All the Worlds a stage, All men are merely players'
cat(str_wrap(some.text, width=25)) # Usage of the str_wrap function.
```

### Extracting Words {.unnumbered}

Let us complete this chapter with the simple **word()** function which extract words from a sentence. We specify the positions of the word to be extracted from the setence. The default separator value is space.

```{r Example 14: extract words}

some.text <- c('The quick brown fox', 'jumps on the brown dog') # Extracting the first two words of a character vector.
word(some.text, start=1, end=2)
word(some.text, start=1, end=-2) # Extracting all but the last word from a character vector.
```

## Regular Expressions {-}

Regular expression is a sequence of characters that describe a patttern to match. The concept was formalized by American mathematician Stephen Cole Kleene. A regular expression pattern can contain a combination of alphanumeric and special characters. Let us take a closer look how these special characters can be used to craft regular expressions in R.

#### Metacharacters {.unnumbered}

Metacharacters are characters that have a special meaning within a regular expression. Unlike other characters that are used to match themselves, metacharacters have a reserved status and cannot be matched explicity. The following table shows a list of metacharacters used in regular expressions.

```{r Table 1: metacharacters, echo=FALSE}
meta <- read.csv(file='tables/metacharacters.csv', header=TRUE)
knitr::kable(
   meta,
  caption='Metacharacters in Regular Expression'
)
```
We shall now see how metacharacters can be used to match different patterns with a few examples.
```{r Example 16: metacharacters, message=FALSE, warning=FALSE}
string <- c('hands', 'data', 'on', 'data$cience', 'handsondata$cience', 'handson')

grep(pattern='^data', string, value=TRUE) # Matching the occurance of pattern at the begining of the string.
grep(pattern='on$', string, value=TRUE) # Matching occurance of pattern at the end of the string.
str_detect(pattern='(nd)+', string) # Detecting if the pattern (nd) occurs atleast ones.
```
Inorder to match a metacharacter in R, we use  $\backslash\backslash$ (double backslash) before them. 
```{r Example 17: matching metacharacter, match_meta}
grep(pattern='\\$', string, value=TRUE) # Matching the metacharacter $ 
```

#### Quantifiers {.unnumbered}

Quantifiers are used to match repitition of pattern within a string. The following table shows a list of quantifiers.

```{r Table 2: quantifiers, echo=FALSE, quantifiers,  echo=FALSE}
quantifier <- read.table(file='tables/quantifier.csv', header=TRUE, sep='\t')
knitr::kable(
   quantifier,
  caption='Quantifiers in Regular Expression'
)
```

Let us see some examples of quantifiers in practice.

```{r Example 18: Quantifiers}
strings <- c('aaab', 'abb', 'bc', 'abbcd', 'bbbc', 'abab', 'caa')
grep(pattern='ab*b', strings, value=TRUE) # Matching strings containining a pattern atleast 0 times.
grep(pattern='abbc?', strings, value=TRUE) # Matching strings containing the pattern atmost ones.
grep(pattern='b{2,}?', strings, value=TRUE) # Matching strings containing the pattern atleast 2 times.
```

#### Character classes {.unnumbered}
A character class is a set that characterises a category of characters. They are enclosed within [] and they match one of the mentioned characters in the set. For example the character class [0-9] matches the first digit occuring in the string. Below are a set of character classes.

```{r Table 3: Character class, echo=FALSE, echo=FALSE, warning=FALSE}
quantifier <- read.table(file='tables/characterclass.txt', header=TRUE, sep='\t')
knitr::kable(
   quantifier,
  caption='Character Class'
)
```

Let us see some simple examples of using character classes in regular expressions.
```{r Example 19: Character class}
string <- c('abc12', '@#$', '345', 'ABcd')
grep(pattern='[0-9]+', string, value=TRUE) # Matching strings that have digits.
grep(pattern='[A-Z]+', string, value=TRUE)# Matching strings that have capital letters.
grep(pattern='[^@#$]+', string, value=TRUE) # Matching strings not do not have special characters.
```

Alternatively R allows the use of POSIX character classes which are represented within [[]] (double braces). 
```{r Example 20: Posix character class}
grep(pattern='[[:alpha:]]', string, value=TRUE) # Matching alpha numeric characters.
grep(pattern='[[:upper:]]', string, value=TRUE) # Matching upper case characters.
```

###Functions in R to Support Regular Expressions {-}

R has a great array of functions that support regular expressions. The following is a list of functions from the base and stringr package that support regular expressions.

```{r Table 4: Functions supporting regular expressions, echo=FALSE}
regex.func <- read.csv('tables/regex_functions.txt', sep=',', header=TRUE)
knitr::kable(
  regex.func,
  caption='Functions Supporting Regular Expressions in R'
)
```

