---
knit: "bookdown::preview_chapter"
---

# Maps and Spatial Visualization

Spatial data is central to many of our tasks as Data Scientists.
Identifying patterns, correlations and relationships between those
patterns delivers opportunities for delivering new services.  Imagine
predicting common routes for travellers this morning, and dynamically
routing public transport to meet those needs.  Fundamental to the Data
Scientist is the ability to process, visualize and then model spatial
data.  Done right maps can be a very effective communications
tool. Numerous R packages work together to bring us a sophisticated
mapping and spatial analysis capability.

```{r loading packages, results='hide', message=FALSE}
library(ggplot2)       # Plotting maps.
library(maps)          # Map data.
library(scales)        # For transperency Functions: alpha() comma()
library(maptools)      # For shapefiles
library(dismo)         # Obtaining gis data
library(broom)         # For conversion functions: tidy() 
library(rgdal)         # For shapefiles
library(dplyr)         # For sample function
library(OpenStreetMap) # OSM maps
library(ggmap)         # Google maps
library(leaflet)       # Interactive Viz
library(magrittr)      # XXXX.
```
```{r, turn_on_cache, echo=FALSE, results='hide'}
knitr::opts_chunk$set(cache=TRUE)
```

##Geocodes {-}

One of the fundamental things about spatial data and mapping is the
geographic coordinate system used to uniquely identify locations. We
use longitude (x axis, abbreviated lon) and latitude (y axis,
abbreviated lat) for locations on our planet.  The longitude is the
angle from the meridian through Greenwich and the latitude is the
angle from the equator.  We can use street addresses and locations.
Here are a few examples.

```{r geocodes, message=FALSE}
geocode("Singapore")
geocode("11 Bunda Street, Canberra")
geocode("Raffles Place", output="more")
```

For later use we will save some of our locations.

```{r geocodes1, message=FALSE}
cityhall <- as.numeric(geocode("City Hall, Singapore")) %T>% print()
raffles <- as.numeric(geocode("Raffles Place")) %T>% print()
sentosa <- as.numeric(geocode("Sentosa")) %T>% print()
```

## World Map {-}

The data for plotting the World map comes from the maps (Brownrigg, 2014) package. We load the vector data for plotting a world map using map_data().
```{r get world map, message=FALSE}
ds <- map_data("world")
head(ds)
```

It is quite simple to plot the world map using ggplot2 (Wickham and Chang, 2014).
```{r world map, message=FALSE}
p <- ggplot(ds, aes(x=long, y=lat, group=group)) + geom_polygon()
p
```

Let us add some colour based on regions. Now there are many regions, so for the same of aesthecity we will turn off the legend.
```{r coloured world map, message=FALSE}

p  <-   ggplot(ds, aes(x=long, y=lat, group=group, fill=region)) +
        geom_polygon() +
        theme(legend.position="none")
p
```

## Understanding Spatial Data {-}

Spatial data involves a combination of information about locations and
geometric features related that best describe the location. The most
commonly used formats to store spatial information are Raster and
Vector.  A Raster format is a type of digital image represented by
grids which can be reduced and enlarged. They are made up of a matrix
of pixels with each pixel representing an area within the spatial
object.  A Vector format consists of geometric shapes organized in the
form of shapefiles. A shapefile consist of three files under the
extensions .shp, .shx and .dbf which contain the features, indices of
shapes and attributes respectively.
      
### Plotting Raster Data{-}
We shall use the getData function in the dismo package to obtain  altitude data of Australia and plot them.
```{r raster plot}
aust <- getData('alt', country='Australia', mask=TRUE)
plot(aust)
```

### Plotting Shapefiles{-}

Let us visualize the streets and places of Singapore. We can obtain the data from <a href="https://data.gov.sg/dataset/street-and-places">data.gov.sg </a>. We shall project it on top of a base map obtained from OpenStreetMap. 


```{r plot shapefile}
shpfile <- readOGR(dsn='data/street-and-places/', 'StreetsandPlaces')
head(coordinates(shpfile), 5)
```

It is observed that the projection system is different than the base map. So we shall transform the projection in to Web Mercator. 
```{r coordinate transform}
crsobj    <- CRS("+proj=longlat +datum=WGS84")   # Web Mercator projection system
shpfile.t <- spTransform(shpfile, crsobj)      # Applying projection transformation
df        <- as.data.frame(coordinates(shpfile.t))   # Converting to a data frame
head(df, 5)
```

Now it looks fine to be projected on top of the base map.

```{r get Singapore map, warning=FALSE}
sgmap <- get_map(location="Singapore", zoom=11,   
                 maptype="roadmap", source="google") # Using Osm base map of Singapore
ggmap(sgmap) +
  geom_point(data=df,
             aes(x=coords.x1, y=coords.x2),
             color='orange', size=1)
```

## Annotating Location on a Map {-}

We shall plot the location of Sentosa and annotate it with a location
landmark and text.

```{r annotation, message=FALSE}
addr <- 'Sentosa' 
loc  <- as.numeric(geocode(addr))
map  <- get_map(loc, zoom=15, maptype='satellite', source='google')
lbl  <- data.frame(lon=loc[1], lat=loc[2], text=addr)

ggmap(map) +
  geom_text(data=lbl, aes(x=lon, y=lat, label=text),
            colour='blue', vjust=1.5) +
  geom_point(data=lbl, aes(x=lon, y=lat),
             colour='red', size=3) +
  xlab('Longitude') + ylab('Latitude')
```

## Plotting Polygons and HeatMaps{-}

Let us plot the male population from the subzone male population <a href="https://data.gov.sg/dataset/singapore-residents-by-subzone-age-group-and-sex-june-2016-male">dataset</a> of Singapore.
We shall first plot the spatial boundaries of the polygon and then overlay the gradients of male population over it.


```{r heatmaps, message=FALSE, warning=FALSE}
sg.popul.spdf <- readOGR(dsn='data/subzone-age-male-2016',
                        'SUBZONE_AGE_MALE_2016', verbose=FALSE)
# Transforming to WSG1984
sg.popul.spdf <- spTransform(sg.popul.spdf, CRSobj=crsobj)
# Converting spatialPolygon to DataFrame
sg.df <- tidy(sg.popul.spdf)
sg.df$id <- as.integer(sg.df$id)
# Creating a dataframe containing the Population data 
sg.data <- data.frame(Population=cut_number(sg.popul.spdf$TOTAL, 3),
           ZONE=sg.popul.spdf$SUBZONE_N,
           id=sg.popul.spdf$OBJECTID)
# Merging the two datasets
sg.df   <- merge(sg.df, sg.data, by='id')
p       <- ggmap(sgmap)
pop.plt <- p + geom_polygon(data=sg.df, aes(x= long, y=lat, fill=Population))
pop.plt <- pop.plt + theme(line=element_blank(),
           axis.title=element_blank(),
           axis.text=element_blank(),
           title=element_text(hjust=0.5)) +
           scale_fill_brewer(palette=1, type='seq') +
           ggtitle('Male Population Singapore 2016') 
pop.plt
```

## Interactive Maps with Leaflet {-}

Earlier we saw a static visualization of places in Singapore now we
can explore an interactive version of the same using leaflet.


```{r interactive maps, leaflet_example, eval=FALSE}
library(leaflet)

smpl <- sample_n(df, size=30)
leaflet() %>%
  addTiles() %>% 
  addMarkers(data=smpl,
             lng=smpl$coords.x1,
             lat=smpl$coords.x2)
```



